### 2015-02-23 doc/thesis/review.tex

Dart \citep{dart} and TypeScript \citep{typescript} are new
languages that are designed with an optional type system.
Both use JavaScript as their code generation target because
their main purpose is web development.
In fact, Dart is a new class-based object-oriented language with
optional type annotations and semantics that resembles the
semantics of Smalltalk, while TypeScript is a strict superset of
JavaScript that provides optional type annotations and class-based
object-oriented programming.
Dart has a nominal type system, while TypeScript has a structural
one, but both are unsound by design.
For instance, Dart has covariant arrays, while TypeScript has
covariant function return types,
besides the interaction between statically and dynamically
typed code that is also unsound.

There is no common formalization for optional type systems, and
each language ends up implementing its optional type system in
its own way.
Strongtalk, Dart, and TypeScript provide an informal description of
their optional type systems rather than a formal one.
In the next section we shall see that we can use some features
of gradual typing \citep{siek2006gradual,siek2007objects} to
formalize optional type systems.

@@

The main goal of gradual typing \citep{siek2006gradual} is to allow
programmers to choose between static and dynamic typing in the same
language.
To do that, \citet{siek2006gradual} extended the simply typed
lambda calculus with the dynamic type $?$, as we shall see in
Figure \ref{fig:gtlc}.
In gradual typing, type annotations are optional, so an untyped
variable is syntactic sugar for a variable whose declared type is
the dynamic type $?$, that is, $\lambda x.e$ is equivalent to $\lambda x{:}?.e$.
Under these circumstances, we view gradual typing as a way to add
a dynamic type to statically typed languages.

\begin{figure}[!ht]
\dstart
$$
\begin{array}{llr}
t ::= & & \emph{types}\\
& \;\; \Number & \emph{base type number}\\
& | \; \String & \emph{base type string}\\
& | \; ? & \emph{dynamic type}\\
& | \; t \rightarrow t & \emph{function types}\\
e ::= & & \emph{expressions}\\
& \;\; l & \emph{literals}\\
& | \; x & \emph{variables}\\
& | \; \lambda x{:}t{.}e & \emph{abstractions}\\
& | \; e_{1} e_{2} & \emph{application}
\end{array}
$$
\dend
\caption{Syntax of the gradually-typed lambda calculus}
\label{fig:gtlc}
\end{figure}

The central idea of gradual typing is the \emph{consistency}
relation, written $t_{1} \sim t_{2}$.
The consistency relation allows implicit conversions to and from the
dynamic type, and disallows conversions between inconsistent types
\citep{siek2006gradual}.
For instance, $\Number \sim \;?$, $? \sim \Number$,
$\String \sim \;?$, and $? \sim \String$,
but $\Number \not\sim \String$, and
$\String \not\sim \Number$.
The consistency relation is both reflexive and symmetric, but
it is neither commutative nor transitive.

\begin{figure}[!ht]
\dstart
$$
\begin{array}{c}
\begin{array}{c}
t \sim t \mylabel{C-REFL}
\end{array}
\;
\begin{array}{c}
t \sim \;? \mylabel{C-DYNR}
\end{array}
\;
\begin{array}{c}
? \sim t \mylabel{C-DYNL}
\end{array}
\\ \\
\begin{array}{c}
\dfrac{t_{3} \sim t_{1} \;\;\; t_{2} \sim t_{4}}
      {t_{1} \rightarrow t_{2} \sim t_{3} \rightarrow t_{4}} \mylabel{C-FUNC}
\end{array}
\end{array}
$$
\dend
\caption{The consistency relation}
\label{fig:consistency}
\end{figure}

@@

Figure \ref{fig:gts} uses the consistency relation in the typing rules
of the gradual type system of the simply typed lambda calculus extended
with the dynamic type $?$.
The environment $\env$ is a function from variables to types, and
the directive $type$ is a function from literal values to types.
The rule \textsc{T-VAR} uses the environment function $\env$ to get the
type of a variable $x$.
The rule \textsc{T-LIT} uses the directive $type$ to get the type of
a literal $l$.
The rule \textsc{T-ABS} evaluates the expression $e$ with an environment
$\env$ that binds the variable $x$ to the type $t_{1}$, and the resulting
type is the the function type $t_{1} \rightarrow t_{2}$.
The rule \textsc{T-APP1} handles function calls where the type of the
function is dynamically typed; in this case, the argument type may have
any type and the resulting type has the dynamic type.
The rule \textsc{T-APP2} handles function calls where the type of the
function is statically typed; in this case, the argument type should
be consistent with the argument type of the function's signature.

\begin{figure}[!ht]
\dstart
$$
\begin{matrix}
\dfrac{\env(x) = t}
      {\env \vdash x:t} \mylabel{T-VAR}
\;\;\;
\dfrac{type(l) = t}
      {\env \vdash l:t} \mylabel{T-LIT}
\\ \\
\dfrac{\env, x:t_{1} \vdash e:t_{2}}
      {\env \vdash \lambda x:t_{1}.e:t_{1} \rightarrow t_{2}} \mylabel{T-ABS}
\;\;\;
\dfrac{\env \vdash e_{1}:\;? \;\;\;
       \env \vdash e_{2}:t}
      {\env \vdash e_{1} e_{2}:\;?} \mylabel{T-APP1}
\\ \\
\dfrac{\env \vdash e_{1}:t_{1} \rightarrow t_{2} \;\;\;
       \env \vdash e_{2}:t_{3} \;\;\;
       t_{3} \sim t_{1}}
      {\env \vdash e_{1} e_{2}:t_{2}} \mylabel{T-APP2}
\end{matrix}
$$
\dend
\caption{Gradual type system gradually-typed lambda calculus}
\label{fig:gts}
\end{figure}

@@

The subtyping relation $\subtype$ is actually a pitfall on Thatte's
quasi-static typing, because it sets the dynamic type
as the top and the bottom of the subtying relation:
$t \subtype \;?$ and $? \subtype t$.
Subtyping is transitive, so we know that
\[
\frac{\Number \subtype \;? \;\;\;
      ? \subtype \String}
     {\Number \subtype \String}
\]
Therefore, downcasts combined with the transitivity of subtyping
accepts programs that should be rejected.

Later, \citet{siek2007objects} reported that the consistency relation is
orthogonal to the subtyping relation, so we can combine them to achieve
the \emph{consistent-subtyping} relation, written $t_{1} \lesssim t_{2}$.
This relation is essential for designing gradual type systems for
object-oriented languages.
Like the consistency relation, and unlike the subtyping relation,
the consistent-subtyping relation is not transitive.
Therefore, $\Number \lesssim \;?$, $? \lesssim \Number$,
$\String \lesssim \;?$, and $? \lesssim \String$,
but $\Number \not\lesssim \String$, and
$\String \not\lesssim \Number$.

@@

\begin{figure}[!ht]
\dstart
$$
\begin{array}{c}
\begin{array}{c}
\Number \subtype \Number \mylabel{S-NUM}
\end{array}
\;
\begin{array}{c}
\String \subtype \String \mylabel{S-STR}
\end{array}
\\ \\
\begin{array}{c}
? \subtype \;? \mylabel{S-ANY}
\end{array}
\;
\begin{array}{c}
\dfrac{t_{3} \subtype t_{1} \;\;\; t_{2} \subtype t_{4}}
      {t_{1} \rightarrow t_{2} \subtype t_{3} \rightarrow t_{4}} \mylabel{S-FUN}
\end{array}
\end{array}
$$
\dend
\caption{The subtyping relation}
\label{fig:subtyping}
\end{figure}

Figure \ref{fig:subtyping} presents the subtyping relation for the simply
typed lambda calculus extended with the dynamic type $?$.
Even though we could have used the reflexive rule $t \subtype t$ to express
the rules \textsc{S-NUM}, \textsc{S-STR}, and \textsc{S-ANY},
we did not combine them into a single rule to make explicit the
neutrality of the dynamic type $?$ to the subtyping rules.
The dynamic type $?$ must be neutral to subtyping to avoid the pitfall
from Thatte's quasi-static typing.
The rule \textsc{S-FUN} defines the subtyping relation for function types,
which are contravariant on the argument type and covariant on the return type.

@@

Still, we can view gradual typing as a way to formalize an optional
type system when the gradual type system does not insert run-time
checks.
BabyJ \citep{anderson2003babyj} and Alore \citep{lehtosalo2011alore}
are two examples of object-oriented languages that have an
optional type system with a formalization that relates to gradual typing,
though the optional type systems of both BabyJ and Alore are nominal.
BabyJ uses the relation $\approx$ that is similar to the consistency
relation while Alore combines subtyping along with the consistency
relation to define a \emph{consistent-or-subtype} relation.
The consistent-or-subtype relation is different from the
consistent-subtyping relation of \citet{siek2007objects}, but it is
also written $t_{1} \lesssim t_{2}$.
The consistent-or-subtype relation holds when $t_{1} \sim t_{2}$
or $t_{1} <: t_{2}$, where $<:$ is transitive and $\sim$ is not.
Alore also extends its optional type system to include optional
monitoring of run-time type errors in the gradual typing style.

>>>>

Dart \citep{dart} and TypeScript \citep{typescript} are new
languages that are designed with an optional type system.
Both use JavaScript as their code generation target because
their main purpose is web development.
In fact, Dart is a new class-based object-oriented language with
optional type annotations and semantics that resembles the
semantics of Smalltalk, while TypeScript is a strict superset of
JavaScript that provides optional type annotations and class-based
object-oriented programming.
Dart has a nominal type system, while TypeScript has a structural
one, but both are unsound by design.
For instance, Dart has covariant arrays, while TypeScript has
covariant parameter types in function signatures,
besides the interaction between statically and dynamically
typed code that is also unsound.

There is no common formalization for optional type systems, and
each language ends up implementing its optional type system in
its own way.
Strongtalk, Dart, and TypeScript provide an informal description of
their optional type systems rather than a formal one.
In the next section we will show that we can use some features
of gradual typing \citep{siek2006gradual,siek2007objects} to
formalize optional type systems.

@@

The main goal of gradual typing \citep{siek2006gradual} is to allow
programmers to choose between static and dynamic typing in the same
language.
To do that, \citet{siek2006gradual} extended the simply typed
lambda calculus with the dynamic type $?$, as we can see in
Figure \ref{fig:gtlc}.
In gradual typing, type annotations are optional, and an untyped
variable is syntactic sugar for a variable whose declared type is
the dynamic type $?$, that is, $\lambda x.e$ is equivalent to $\lambda x{:}?.e$.
Under these circumstances, we view gradual typing as a way to add
a dynamic type to statically typed languages.

\begin{figure}[!ht]
\dstart
$$
\begin{array}{llr}
T ::= & & \textsc{types}\\
& \;\; \Number & \textit{base type number}\\
& | \; \String & \textit{base type string}\\
& | \; ? & \textit{dynamic type}\\
& | \; T \rightarrow T & \textit{function types}\\
e ::= & & \textsc{expressions}\\
& \;\; l & \textit{literals}\\
& | \; x & \textit{variables}\\
& | \; \lambda x{:}T{.}e & \textit{abstractions}\\
& | \; e_{1} e_{2} & \textit{application}
\end{array}
$$
\dend
\caption{Syntax of the gradually-typed lambda calculus}
\label{fig:gtlc}
\end{figure}

The central idea of gradual typing is the \emph{consistency}
relation, written $T_{1} \sim T_{2}$.
The consistency relation allows implicit conversions to and from the
dynamic type, and disallows conversions between inconsistent types
\citep{siek2006gradual}.
For instance, $\Number \sim \;?$, $? \sim \Number$,
$\String \sim \;?$, and $? \sim \String$,
but $\Number \not\sim \String$, and
$\String \not\sim \Number$.
The consistency relation is both reflexive and symmetric, but
it is neither commutative nor transitive.

\begin{figure}[!ht]
\dstart
$$
\begin{array}{c}
\begin{array}{c}
T \sim T \mylabel{C-REFL}
\end{array}
\;
\begin{array}{c}
T \sim \;? \mylabel{C-DYNR}
\end{array}
\;
\begin{array}{c}
? \sim T \mylabel{C-DYNL}
\end{array}
\\ \\
\begin{array}{c}
\dfrac{T_{3} \sim T_{1} \;\;\; T_{2} \sim T_{4}}
      {T_{1} \rightarrow T_{2} \sim T_{3} \rightarrow T_{4}} \mylabel{C-FUNC}
\end{array}
\end{array}
$$
\dend
\caption{The consistency relation}
\label{fig:consistency}
\end{figure}

@@

Figure \ref{fig:gts} uses the consistency relation in the typing rules
of the gradual type system of the simply typed lambda calculus extended
with the dynamic type $?$.
The environment $\env$ is a function from variables to types, and
the directive $type$ is a function from literal values to types.
The rule \textsc{T-VAR} uses the environment function $\env$ to get the
type of a variable $x$.
The rule \textsc{T-LIT} uses the directive $type$ to get the type of
a literal $l$.
The rule \textsc{T-ABS} evaluates the expression $e$ with an environment
$\env$ that binds the variable $x$ to the type $T_{1}$, and the resulting
type is the the function type $T_{1} \rightarrow T_{2}$.
The rule \textsc{T-APP1} handles function calls where the type of a
function is dynamically typed; in this case, the argument type may have
any type and the resulting type has the dynamic type.
The rule \textsc{T-APP2} handles function calls where the type of a
function is statically typed; in this case, the argument type should
be consistent with the argument type of the function's signature.

\begin{figure}[!ht]
\dstart
$$
\begin{matrix}
\dfrac{\env(x) = T}
      {\env \vdash x:T} \mylabel{T-VAR}
\;\;\;
\dfrac{type(l) = T}
      {\env \vdash l:T} \mylabel{T-LIT}
\\ \\
\dfrac{\env[x \mapsto T_{1}] \vdash e:T_{2}}
      {\env \vdash \lambda x:T_{1}.e:T_{1} \rightarrow T_{2}} \mylabel{T-ABS}
\;\;\;
\dfrac{\env \vdash e_{1}:\;? \;\;\;
       \env \vdash e_{2}:T}
      {\env \vdash e_{1} e_{2}:\;?} \mylabel{T-APP1}
\\ \\
\dfrac{\env \vdash e_{1}:T_{1} \rightarrow T_{2} \;\;\;
       \env \vdash e_{2}:T_{3} \;\;\;
       T_{3} \sim T_{1}}
      {\env \vdash e_{1} e_{2}:T_{2}} \mylabel{T-APP2}
\end{matrix}
$$
\dend
\caption{Gradual type system gradually-typed lambda calculus}
\label{fig:gts}
\end{figure}

@@

The subtyping relation $\subtype$ is actually a pitfall on Thatte's
quasi-static typing, because it sets the dynamic type
as the top and the bottom of the subtying relation:
$T \subtype \;?$ and $? \subtype T$.
Subtyping is transitive, so we know that
\[
\frac{\Number \subtype \;? \;\;\;
      ? \subtype \String}
     {\Number \subtype \String}
\]
Therefore, downcasts combined with the transitivity of subtyping
accepts programs that should be rejected.

Later, \citet{siek2007objects} reported that the consistency relation is
orthogonal to the subtyping relation, so we can combine them to achieve
the \emph{consistent-subtyping} relation, written $T_{1} \lesssim T_{2}$.
This relation is essential for designing gradual type systems for
object-oriented languages.
Like the consistency relation, and unlike the subtyping relation,
the consistent-subtyping relation is not transitive.
Therefore, $\Number \lesssim \;?$, $? \lesssim \Number$,
$\String \lesssim \;?$, and $? \lesssim \String$,
but $\Number \not\lesssim \String$, and
$\String \not\lesssim \Number$.

@@

\begin{figure}[!ht]
\dstart
$$
\begin{array}{c}
\begin{array}{c}
\Number \subtype \Number \mylabel{S-NUM}
\end{array}
\;
\begin{array}{c}
\String \subtype \String \mylabel{S-STR}
\end{array}
\\ \\
\begin{array}{c}
? \subtype \;? \mylabel{S-ANY}
\end{array}
\;
\begin{array}{c}
\dfrac{T_{3} \subtype T_{1} \;\;\; T_{2} \subtype T_{4}}
      {T_{1} \rightarrow T_{2} \subtype T_{3} \rightarrow T_{4}} \mylabel{S-FUN}
\end{array}
\end{array}
$$
\dend
\caption{The subtyping relation}
\label{fig:subtyping}
\end{figure}

Figure \ref{fig:subtyping} presents the subtyping relation for the simply
typed lambda calculus extended with the dynamic type $?$.
Even though we could have used the reflexive rule $T \subtype T$ to express
the rules \textsc{S-NUM}, \textsc{S-STR}, and \textsc{S-ANY},
we did not combine them into a single rule to make explicit the
neutrality of the dynamic type $?$ to the subtyping rules.
The dynamic type $?$ must be neutral to subtyping to avoid the pitfall
from Thatte's quasi-static typing.
The rule \textsc{S-FUN} defines the subtyping relation for function types,
which are contravariant on the argument type and covariant on the return type.

@@

Still, we can view gradual typing as a way to formalize an optional
type system when the gradual type system does not insert run-time
checks.
BabyJ \citep{anderson2003babyj} and Alore \citep{lehtosalo2011alore}
are two examples of object-oriented languages that have an
optional type system with a formalization that relates to gradual typing,
though the optional type systems of both BabyJ and Alore are nominal.
BabyJ uses the relation $\approx$ that is similar to the consistency
relation while Alore combines subtyping along with the consistency
relation to define a \emph{consistent-or-subtype} relation.
The consistent-or-subtype relation is different from the
consistent-subtyping relation of \citet{siek2007objects}, but it is
also written $T_{1} \lesssim T_{2}$.
The consistent-or-subtype relation holds when $T_{1} \sim T_{2}$
or $T_{1} <: T_{2}$, where $<:$ is transitive and $\sim$ is not.
Alore also extends its optional type system to include optional
monitoring of run-time type errors in the gradual typing style.
